Counterexamples in Mathematical Relations

### 2024-11-07 Relations/notes.tex
created notes.tex to merge various scattered bullet points

\textbf{Primary Goal: A technical report that focuses on
Abstract Rewriting in Constructive Setting }

\section{2024.10.31}

Rewriting properties.
Elementwise
\begin{itemize}
  \item NF
  \item WN
  \item SN
  \item WCR
  \item CR
  \item NFP
  \item UN$\to$
  \item UN
  \item RP
  \item RP$-$
  \item $\omega$-bound
  \item Recurrent
  \item CP (Cofinality property)
  \item FB (Finite branching of rewrite steps)
  \item DominatedByWF (Needed?)
  \item Define WNFP to be:
  \[\forall a b c \to b \in NF \to a R^* b \to a R^* c \to c R^* b \]
  \item Does WNFP and WN imply CR?  Looks like it!
  \item Another interesting one? $WN^*$: hereditarily $WN$ ($WN$ for the reduction graph starting from a given element $x$.)

@@

Other properties that show up in classical ARS theory:
\begin{itemize}
  \item Inc
  \item Ind
  \item $\mathrm{CR}^{\le 1}$
\end{itemize}
Want to discuss these, and why we may or may not include them in our
development.

Outline of the document:
\begin{enumerate}
  \item Create a diagram with implications between these, and
  combinations of these, similar to the one in TeReSe.
  \item Formalization of valid implications
  \item Counterexamples: Compile a list/chart/graph of those relationships
  which can be disproved. (A big part!!)
  \item A list of classical properties, and elaboration of their meaning
  and whether it's reasonable to assume them in a rewriting context.
  (Discuss each, and its role in ARS theory.)
  \item Well-foundedness: different constructive formulations, and relationships between them
  \item ``Guide'' to the formalization code, GitHub repo
  \item Discussion of open problems (Add a subsection of these at the end of each section?)
  \item Conclusion
\end{enumerate}

\section{Established implications:}

@@

\section{2024.10.10}

\begin{itemize}
  \item Look at restricting WFmin to $\lnot\lnot$-closed predicates
  \item Status of important hypotheses:
    \begin{itemize}
      \item Deciding whether a given element $x$ reduces to a given element $y$
      (Is $R$ itself decidable?)
      \item Deciding whether a given element $x$ reduces to some element $y$
      (Is being an $R$-normal form decidable?)

      \item Does UN-lemma \emph{require} the decidability assumption?
    \end{itemize}
  \item META questions: Which of the ARS theorems can be made ``local/pointwise'',
  for example,
  \begin{itemize}
    \item NL+: $WCR(R) \land SN(x) \to CR(x)$?  (Yes?)
    \item NL++: $WCR(x) \land SN(x) \to CR(x)$? (No.)
  \end{itemize}
  \item When time permits, uniformize the notation (variable names, etc.) to the extent possible.
\end{itemize}

\section{2024.09.19}

TODO item: Compile a list of counterexamples

\section{2024.09.12}
\begin{itemize}
  \item Investigate \texttt{iii-lemma}.  (Is it false, or is there a classical proof?)
  (No constructive proof?)
  \item Look at variations of RP.
  \item What happens if properties involving sequences are reformulated for
  $R^r$? (Cannot decide whether the given sequence is finite or infinite.)
  \item Look at RP- and WCR implies RP!
  \item Can "sequential" properties like RP, RP-, isWFseq, isWFseq-, $\omega$-bounded, etc.,
  be related to the cofinality property.
\end{itemize}

\section{Relations, ARS}

\begin{enumerate}
  \item Define logical and structurual operations on relations, proving laws about them.
  \item Define closure operations on realtions, proving laws about them
  \item Define properties such as transtivity etc.
\end{enumerate}

\subsection{Well-foundedness}

Four notions:
\begin{itemize}
  \item Based on every element being accessible
  \item Every inductive relation is universally true
  \item Every non-empty set has a minimal element
  \item No infinite sequence is decreasing
\end{itemize}

\section{Open research questions}
\begin{enumerate}
  \item Sufficient conditions for reversing the implications between notions of
  well-foundedness?
  \item Knaster-Tarski Lemma: Most general form? Alternative formulations?
  \item Syntax for closure operators allowing to prove the needed properties
  uniformly/generically
\end{enumerate}

\section{To Do}
\begin{enumerate}
  \item Clean up the current development
  \item More laws about closure operations: Monotonicity, idempotency
  \item Refactor the ARS formalized results with the new development
  \item Knaster--Tarski in a well-founded setting: what additional
  hypotheses are necessary?
\end{enumerate}

\subsection{Questions about well-foundedness}

\begin{enumerate}
  \item For a given element $x$, is the property of
  being accessible $\lnot\lnot$-closed? I.E.,
  Does $x$ being $\lnot\lnot$-accessible imply that $x$ is accessible?
  (Conjecture: No.)

@@

   \item (2024.10.10) Is being $R$-WN inductive?
   \item (2024.10.10) Does $R$ being finitely branching imply
   that, for an inductive $\phi$, if $\lnot \phi(x)$ then
   $\exists y. Rxy \land \lnot \phi(y)$?

@@

\section{Open problems}

\begin{itemize}
  \item The infinite haircomb refutes $WN \land RP \to \omega$-bounded.  But it is not WCR.

@@

  \item RP- and $\omega$-bounded implies RP?
  \item RP- and WCR imply RP?
  \item
  \[ WNg \land UN \to CRelem :
  \forall (R : \mathscr{R} A) \to WN R \to \forall x \to is R -UN x \to is R -CR x \]
  \item $WN R \to UN R \to \omega-bdd R \to SN R$ (1.2.3.ii-)
\end{itemize}

>>>>

This document aims to bring together the various notes from report, rewriting qs, and counterexamples.

\section{Definitions and implications}
Rewriting properties.
Elementwise
\begin{itemize}
  \item NF
  \item WN
  \item SN
  \item WCR
  \item CR
  \item NFP
  \item UN$\to$
  \item UN
  \item RP
  \item RP$-$
  \item $\omega$-bound
  \item Recurrent
  \item CP (Cofinality property)
  \item FB (Finite branching of rewrite steps)
  \item DominatedByWF (Needed?)
  \item Define WNFP to be:
  \[\forall a b c \to b \in NF \to a R^* b \to a R^* c \to c R^* b \]
  \item Does WNFP and WN imply CR?  Looks like it!
  \item Another interesting one? $WN^*$: hereditarily $WN$ ($WN$ for the reduction graph starting from a given element $x$.)

@@

Other properties that show up in classical ARS theory:
\begin{itemize}
  \item Inc
  \item Ind
  \item $\mathrm{CR}^{\le 1}$
\end{itemize}
Want to discuss these, and why we may or may not include them in our
development.
\newpage
\section{Formalization of implications}
\section{Established implications:}

@@

\newpage
\section{Counterexamples}
\subsection{Counterexample 1}
\begin{center}

    \begin{tikzcd}[node distance=2cm, auto]
        % Nodes
        \node (a) {a};
        \node (b) [right of=a, xshift=1cm] {b};
        \node (c) [right of=b, xshift=1cm] {c};
        \node (d) [right of=c, xshift=1cm] {d};

        % Arrows
        \draw[->] (b) to [bend left] (c);
        \draw[->] (c) to [bend left] (b);
        \draw[->] (b) -- (a);
        \draw[->] (c) -- (d);
    \end{tikzcd}

\end{center}
Counterexample to:
\begin{itemize}
    \item WCR $\to$ CR
    \item WCR and WN $\to$ CR
    \item WCR and WN $\to$ SN
    \item \mbox{WCR R $\to$ $\forall$ a b $\to$ is R -NF b $\to$ (R *) a b $\to$ $\forall$ c $\to$ (R *) a c $\to$ (R *) c b}
    \item IOW, WCR $\to$ WNFP
\end{itemize}

\subsection{Counterexample 2}
\begin{center}
    \begin{tikzcd}
        % Nodes
        \node (x) {x};
        \node (n) [below left of=x, xshift=-1cm, yshift=-1cm] {n \in NF};
        \node (y) [below right of=x, xshift=1cm, yshift=-1cm] {y};
        \node (z) [below of=y, yshift=-1cm] {z};

        % Arrows
        \draw[->>] (x) -- (n);
        \draw[->>] (x) -- (y);
        \draw[->] (y) to [bend left] (z);
        \draw[->] (z) to [bend left] (y);
    \end{tikzcd}
\end{center}

Counterexample to:
\begin{itemize}
    \item WN$\downarrow\subseteq$WN :  $\forall$ x $\to$ is R-WN x $\to$ $\forall$ y $\to$ (R *) x y $\to$ is R-WN y
    \item WN$\downarrow$UN$\rightarrow\subseteq$WN :

    UN$\rightarrow\,$R $\to \forall$ x $\to$ is R-WN x $\to$ $\forall$ y $\to$ (R *) x y $\to$ is R-WN y

    \textcolor{red}{SA: The above is a bit messy, can we make it cleaner?}
    \item $\forall$ (R : $\mathscr{R}$ A) $\to$ $\forall$ is R -WN x $\to$ is R -UN x $\to$ is R -CR x

\end{itemize}

\subsection{Counterexample 3}
\begin{center}
  \begin{tikzpicture}
      % Define the nodes inside the region f
      \node (f0) at (0, 0) {$f_0$};
      \node (f1) at (1, 0) {$f_1$};
      \node (fi) at (3, 0) {$f_i$};
      \node (fi+1) at (4, 0) {$f_{i+1}$};
      \node (dots) at (2, 0) {$...$};
      \node (dots2) at (5, 0) {$...$};

      % Draw the arrows between the nodes inside the region f
      \draw[->] (f0) -- (f1);
      \draw[->] (f1) -- (dots);
      \draw[->] (dots) -- (fi);
      \draw[->] (fi) -- (fi+1);
      \draw[->] (fi+1) -- (dots2);

      % Define the region f
      \node[draw, ellipse, fit=(f0) (dots2), label=left:$f$] {};

      % Define the nodes outside the region f
      \node (a) [below=of fi, yshift=-.5cm] {\parbox{2cm}{\centering $a$ \\ $\omega-bound$}};
      \node (y) [above right=of fi, yshift=.5cm] {$y$};

          % Draw the arrows between the nodes inside and outside the region f
          \draw[->>] (fi) to [bend right=15] (a);
          \draw[->>] (a) to [bend right=15] (fi);
          \draw[->>] (fi+1) -- (a);
          \draw[->>] (dots2) -- (a);
          \draw[->>] (fi) -- (y);   
      \end{tikzpicture}
    \end{center}
    Counterexample to:
    \begin{itemize}
      \item ***** DOES NOT WORK ******
      \item RP-$\to$RP 
      \item RP-$\land$WCR$\to$RP \textcolor{red}{NO< NEEDS A NEW CE}
\end{itemize}

\subsection{Other questions}
\begin{enumerate}
  \item If WCR holds, is WN$\downarrow\subseteq$WN ?
  \item Same question for next bullet in 2.2
  \item Same for bullet 3 in 2.2 (but pointwise!)
  \item Do RP- $\land$ WCR imply RP?
\end{enumerate}
\subsection{ARS in Agda}

Important counterexamples:
\begin{enumerate}
  \item $a,b,c,d$ with sinks $a$ and $d$ and a 2-cycle between $b$ and $c$.
  Counterexample to:
  \begin{itemize}
      \item WCR and WN implies CR.
  \end{itemize}
  \item Loop $a \to a$.
  Counterexample to:
  \begin{itemize}
    \item $\omega$-bounded and RP implies SN
    \item $\omega$-bounded and RP implies WN
  \end{itemize}
  \item Infinite sequence $a_0 \to a_1 \to \cdots$. Falsifies:
  \begin{itemize}
    \item $WN$
    \item $\omega$-bounded
    \item RP $\to$ $\omega$-bounded
    \item $WCR \to WN$
    \item $CR \to WN$
    \item $CR, CP \to WN$
    \item $CR, CP \to \omega$-bounded
  \end{itemize}
  \item Infinite sequence $a_0 \to a_1 \to \cdots$ with a colimit $a_\infty$,
  with each $a_n \to a_\infty$ and $a_\infty$ a normal form.

  Counterexample to:
  \begin{itemize}
    \item WN $\to$ SN
    \item WN, CR $\to$ SN
    \item WN, CR, $\omega$-bounded $\to$ SN
  \end{itemize}

  \item Infinite sequence $a_0 \to a_1 \to \cdots$, together with links
  $a_n \to a_0$, for each $n$.
  \item The combination of the last two.
  Counterexample to
  \begin{itemize}
    \item RP
  \end{itemize}
  \item An ``infinite haircomb''. Infinite sequence $a_0 \to a_1 \to \cdots$, together with
  a disjoint, discrete set $b_0, b_1, \dots$, and links $a_n \to b_n$.
  Counterexample to:
  \begin{itemize}
    \item $WN\to WCR$
    \item $WN\to \omega-bounded$
    \item $WN \to SN$
    \item $WN,RP \to SN$
  \end{itemize}
  \item A variation of the above with backlinks $a_n \to a_0$ for all $n$.
\end{enumerate}
\newpage
\newpage
\section{Classical properties}
Principles we are considering
\begin{itemize}
  \item Decidability of equality on $A$
  \item Decidability of the relation $R$
  \item When relevant, decidability of the given precicate $P$
  \item Not-not-closure of being accessible
  \item Not-not-closure of $R$
  \item Not-not-closure of $P$
  \item Decidability of whether a given element is an $R$-normal form
  \item When relevant, local/element-wise versions of the above
  \item Markov's principle:
  \begin{itemize}
    \item In $wfMin_0 \to wfSeq$.

    Given a sequence $s : \nat \to A$, let $E(a) = \exists n \in \nat. s n \equiv a$.

    Need: $\forall a. \lnot\lnot E(a) \to E(a)$, i.e.,
    \[\lnot\lnot \big(\exists n \in \nat. s n \equiv a\big)
          \to \big(\exists n \in \nat. s n \equiv a\big)\]

  \end{itemize}
\end{itemize}
\newpage
\section{Well foundedness }
\begin{enumerate}
    \item For a given element $x$, is the property of
    being accessible $\lnot\lnot$-closed? I.E.,
    Does $x$ being $\lnot\lnot$-accessible imply that $x$ is accessible?
    (Conjecture: No.)

@@

@@

\newpage
\section{Guide to the code}
\section{Open problems}
\begin{itemize}
    \item Look at restricting WFmin to $\lnot\lnot$-closed predicates
    \item Status of important hypotheses:
      \begin{itemize}
        \item Deciding whether a given element $x$ reduces to a given element $y$
        (Is $R$ itself decidable?)
        \item Deciding whether a given element $x$ reduces to some element $y$
        (Is being an $R$-normal form decidable?)

        \item Does UN-lemma \emph{require} the decidability assumption?
      \end{itemize}
    \item META questions: Which of the ARS theorems can be made ``local/pointwise'',
    for example,
    \begin{itemize}
      \item NL+: $WCR(R) \land SN(x) \to CR(x)$?  (Yes?)
      \item NL++: $WCR(x) \land SN(x) \to CR(x)$? (No.)
    \end{itemize}
    \item When time permits, uniformize the notation (variable names, etc.) to the extent possible.
    \item Can "sequential" properties like RP, RP-, isWFseq, isWFseq-, $\omega$-bounded, etc.,
    be related to the cofinality property.
    \item What happens if properties involving sequences are reformulated for
  $R^r$? (Cannot decide whether the given sequence is finite or infinite.)
  \item Knaster-Tarski Lemma: Most general form? Alternative formulations?
  \item Syntax for closure operators allowing to prove the needed properties
  uniformly/generically
  \item More laws about closure operations: Monotonicity, idempotency
  \item \item Knaster--Tarski in a well-founded setting: what additional
  hypotheses are necessary?
  \item (2024.10.10) Is being $R$-WN inductive?
    \item (2024.10.10) Does $R$ being finitely branching imply
     that, for an inductive $\phi$, if $\lnot \phi(x)$ then
     $\exists y. Rxy \land \lnot \phi(y)$?

     \item The infinite haircomb refutes $WN \land RP \to \omega$-bounded.  But it is not WCR.

@@

     \item RP- and $\omega$-bounded implies RP?
     \item RP- and WCR imply RP?
     \item
     \[ WNg \land UN \to CRelem :
     \forall (R : \mathscr{R} A) \to WN R \to \forall x \to is R -UN x \to is R -CR x \]
     \item $WN R \to UN R \to \omega-bdd R \to SN R$ (1.2.3.ii-)
      \item Finish formulating the ``compactness property".  Two candidates:
       \item ``Every cocone for a given infinite sequence loops back to some point in the sequence"
       \item ``If an $R^*$-sequence has a cocone then it's constant after some point on.''
       \item is $\omega$-bounded implied by WN and the following confluence property:
       $a \to^* n \in NF$, and $a \to b$ then $b \to^* n $.
    \item  Make the first argument to "being a normal form" implicit?
    \item Write up the classical proof to theorem 1-2-3 iii

\end{itemize}
\section{Conclusions}
